The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-05-12 paper/paper.Rmd
Moved code to Appendix

We first manually define an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019].^[In the Supporting Information we present an example where the folds themselves are created automatically, as opposed to manually as in the example.] This `foldID` variable will be used in Section \ref{spatial-cross-validation}.

@@

This process is repeated for each of the four folds acting as the test set, then the four RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values. Furthermore, `comp_dist` can now be specified as a vector of competitive distances to consider.

@@

# Conclusion and future work

@@

>>>>

We first manually define an `sf` object defining four folds that partition the study region. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct `foldID` [@valavi_blockcv_2019]. This `foldID` variable will be used in Section \ref{spatial-cross-validation}.

@@

This process is repeated for each of the four folds acting as the test set, then the four RMSE's are averaged to provide a single estimate of model error. This algorithm is implemented in `run_cv()`, which acts as a wrapper function to both `comp_bayes_lm()` that fits the model and `predict()` that returns predicted values.

@@

# Conclusion

@@

# Appendix

This code replicates Figure \ref{fig:scbi-simulation}: A comparison of root mean squared error of
models for standard, permuted, and spatial cross- validated error estimates.

```{r, eval = FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(viridis)
library(forestecology)
library(blockCV)
library(tictoc)

# Compute growth of trees based on census data -------------------------
census_2013_scbi <- here("paper/scbi.stem2.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- here("paper/scbi.stem3.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  )

# Add spatial information ----------------------------------------------
# Define buffer region using competitive distance range
comp_dist <- 7.5

study_region_scbi <- tibble(
  x = c(0, 300, 300, 0, 0),
  y = c(300, 300, 600, 600, 300)
) %>%
  sf_polygon()

growth_scbi <- growth_scbi %>%
  add_buffer_variable(size = comp_dist, region = study_region_scbi)

# Manually define spatial blocks to act as folds
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 450), c(0, 450))
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 450), c(150, 450))
fold3 <- rbind(c(0, 450), c(150, 450), c(150, 600), c(0, 600))
fold4 <- rbind(c(150, 450), c(300, 450), c(300, 600), c(150, 600))
n_fold <- 4

# Associate each observation to a fold
SpatialBlock_scbi <- spatialBlock(
  speciesData = growth_scbi, k = n_fold, selection = "systematic",
  blocks = blocks_scbi, showBlocks = FALSE, verbose = FALSE
)

growth_scbi <- growth_scbi %>%
  mutate(foldID = SpatialBlock_scbi$foldID %>% factor())

# Compute focal versus competitor tree information ---------------------
focal_vs_comp_scbi <- growth_scbi %>%
  create_focal_vs_comp(comp_dist, blocks = blocks_scbi, id = "stemID")

# Fit model and make predictions ---------------------------------------
# Number of permutation shuffles:
num_shuffle <- 49

# Save results here
run_time <- 0
observed_RMSE <- 0
observed_RMSE_CV <- 0
shuffle_RMSE <- vector("list", 1)
shuffle_RMSE_CV <- vector("list", 1)
filename <- here("paper/simulation_results/") %>%
  str_c("2021-03-03_scbi_", num_shuffle, "_shuffles")

# Run all simulations
# 0. Setup simulation for this species type ----
# Start clock
tic()

# 1. Compute observed test statistic: RMSE with no cross-validation ----
# Fit model (compute posterior parameters)
comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = FALSE)

# Make predictions and compute RMSE
observed_RMSE <- focal_vs_comp_scbi %>%
  mutate(growth_hat = 
           predict(comp_bayes_lm_scbi, focal_vs_comp_scbi)) %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)

# 2. Compute observed test statistic: RMSE with cross-validation -------
observed_RMSE_CV <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, blocks = blocks_scbi) %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)

# 3. Permutation distribution: RMSE with no cross-validation -----------
# Compute num_shuffle permutation test statistics
shuffle_RMSE <- numeric(length = num_shuffle)
for(j in 1:num_shuffle){
  # Fit model (compute posterior parameters) with shuffling
  comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
    comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)

  # Make predictions and compute RMSE
  shuffle_RMSE[j] <-  focal_vs_comp_scbi %>%
    mutate(growth_hat = 
             predict(comp_bayes_lm_scbi, focal_vs_comp_scbi)) %>%
    rmse(truth = growth, estimate = growth_hat) %>%
    pull(.estimate)
}

# 4. Permutation distribution: RMSE with cross-validation --------------
# Compute num_shuffle permutation test statistics
shuffle_RMSE_CV <- numeric(length = num_shuffle)

# Compute num_shuffle permutation test statistics
for(j in 1:num_shuffle){
  # Compute and save RMSE
  shuffle_RMSE_CV[j] <- focal_vs_comp_scbi %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi, 
           run_shuffle = TRUE) %>%
    rmse(truth = growth, estimate = growth_hat) %>%
    pull(.estimate)

  # Status update
  str_c("Shuffle with permutation ", j, " at ", Sys.time()) %>% print()
}

# 5. Save results ----
clock <- toc(quiet = TRUE)
run_time <- clock$toc - clock$tic

model_comp_tbl <- tibble(
  run_time = run_time,
  observed_RMSE = observed_RMSE,
  observed_RMSE_CV = observed_RMSE_CV,
  shuffle_RMSE = shuffle_RMSE,
  shuffle_RMSE_CV = shuffle_RMSE_CV,
)
save(model_comp_tbl, file = filename %>% str_c(".RData"))

# Visualize results ----------------------------------------------------
model_comp <- bind_rows(
  model_comp_tbl %>% 
    select(run_time, observed = observed_RMSE, 
           shuffle = shuffle_RMSE) %>% 
    mutate(CV = FALSE),
  model_comp_tbl %>% 
    select(run_time, observed = observed_RMSE_CV, 
           shuffle = shuffle_RMSE_CV) %>% 
    mutate(CV = TRUE)
) %>%
  gather(type, RMSE, -c(run_time, CV))

model_comp_observed <- model_comp %>%
  filter(type == "observed") %>%
  unnest(cols = c(RMSE))
model_comp_shuffle <- model_comp %>%
  filter(type == "shuffle") %>%
  unnest(cols = c(RMSE))

cv_plot <- ggplot() +
  geom_vline(data = model_comp_observed, 
             aes(xintercept = RMSE, col = CV), 
             linetype = "dashed", show.legend = F) +
  geom_histogram(data = model_comp_shuffle, 
                 aes(x = RMSE, fill = CV), bins = 50) +
  labs(
    fill = "Cross-validated?",
    x = expression(paste("RMSE (cm ", y^{-1}, ")"))
  ) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  scale_fill_viridis(discrete = TRUE) +
  theme_light()
cv_plot

filename %>%
  str_c(".pdf") %>%
  ggsave(plot = cv_plot, width = 16/2, height = 9/2)
```
